\chapter{Making It Happen}
\vspace{-.53in}
   \noindent\color{graylight}\rule[0cm]{3.25in}{0.03cm} \\
    \noindent\color{graylight}\rule[0.4cm]{3.25in}{0.03cm} \\
\color{black}
\vspace{.05in}

\newthought{Statistics education} has come a long way since Fisher and Snedecor \added{XX add citation in margin?}. Moreover, teachers of statistics across the country have generally been enthusiastic about adopting modern methods and approaches. Nevertheless, changing the way we teach isn't always easy. In a way, we are all teachers and learners, a bit like hermit crabs:  To grow, we must first abandon the protective shell of what we are used to and endure a period of vulnerability until we can settle into a new and larger set of habits and expectations. 
 
We have presented many ideas in this report.  We advise readers to move in the directions suggested by taking small steps at first.  Examples of small steps include the following:

\renewcommand{\labelitemi}{$\filledsquare$}

\begin{itemize} [leftmargin=1cm, itemsep=.2em]
\item Adding an activity to your course
\item Having your students do a small project
\item Integrating an applet into a lecture
\item Demonstrating the use of software to your students
\item Increasing the use of real data sets
\item Deleting a topic from the list you currently try to cover to focus more on understanding concepts
\end{itemize}

Your teaching philosophy will inform your choice of textbook, but the recommendations in this report are not about choosing a text.  They are about a way of teaching. 
 
There are many resources available, including the MAA Notes volumes that deal with teaching statistics, the Consortium to Advance Undergraduate Statistics Education (CAUSE) (causeweb.org), the Isostat discussion list (\url{www.lawrence.edu/fac/jordanj/isostat.html}), the SIGMAA- Stat Ed group within the MAA (\url{www.pasles.org/sigmaastat}), and the ASA website, especially the Center for Statistics Education (\url{www.amstat.org/education}) and the Statistical Education Section (\url{www.amstat.org/sections/educ}).
 
\section{\textbf{GAISEing into the Future}}
 
\newthought{A good deal of progress has been made}, but there is still plenty of room to improve the introductory statistics course. Moreover, this course must be flexible and adapt to change as more students enter college having learned aspects of statistics in elementary and secondary school. The Advanced Placement course continues to change the statistics education landscape. Although we have been addressing the general introductory course, we must be mindful of other courses, such as business statistics and mathematical statistics, and of the content and goals of good second courses in statistics that build on the solid conceptual understanding developed in the first course.


\part*{\allcaps{Appendices}}

\noindent \smallcaps{Examples and commentary in these appendices are provided for additional guidance, clarification, and illustration of the guidelines in the main report.} \\
%\vspace{-.53in}
   \noindent\color{graylight}\rule[0cm]{3.25in}{0.03cm} \\
    \noindent\color{graylight}\rule[0.4cm]{3.25in}{0.03cm} \\
\color{black}
\vspace{.05in}

%Redefine the first level
\renewcommand{\theenumi}{\arabic{enumi}.}
\renewcommand{\labelenumi}{\theenumi}
%Redefine the second level
\renewcommand{\theenumii}{\alph{enumii}.}
\renewcommand{\labelenumii}{\theenumii}

\begin{description}
  \item[\allcaps{\textbf{Appendix A: Examples of Activities and Projects}}]\hfill
  \vspace{-.2in}
	\begin{enumerate}[leftmargin=1cm, itemsep=.2em]
	\item Desirable Characteristics of Class Activities
	\item Activities That Could Be Improved
		\begin{enumerate}[leftmargin=1cm, itemsep=.2em]
		\item Pepsi vs. Coke Activity
		\item A Central Limit Theorem Activity
		\end{enumerate}
	\item Additional Examples of Activities and Projects
		\begin{enumerate}[leftmargin=1cm, itemsep=.2em]
		\item Data Gathering and Analysis:  A Class of Projects
		\item Team constructed questions about relationships
		\item Comparing Manual Dexterity under Two Conditions\\
		\end{enumerate}
	\end{enumerate}
 
 \vspace{.25in}
\item[\allcaps{\textbf{Appendix B: Examples of Assessment Items}}] \hfill
\vspace{-.2in}
	\begin{enumerate}[leftmargin=1cm, itemsep=.2em]
	\item Examples of Assessment Items with Problems and Commentary About the Nature of the Difficulty
	\item Examples Showing Ways to improve Assessment Items
	\item Additional Examples of Good Assessment Items\\
	\end{enumerate}
  
  \vspace{.25in}
\item[\allcaps{\textbf{Appendix C: Example of Using Technology}}] \hfill \\
\vspace{-.1in}
A Technology-Based Simulation to Examine the Effectiveness of Treatments for Cocaine Addiction\\

\vspace{.3in}
\item[\allcaps{\textbf{Appendix D: Examples of Naked, Realistic, and Real Data}}] \hfill
	\vspace{-.2in}
	\begin{enumerate}[leftmargin=1cm, itemsep=.2em]
	\item Naked Data (not recommended)
	\item Realistic Data (better, but still not the best)
	\item Real Data (recommended)
	\end{enumerate} 
\end{description}


